\bf Visual Inference for a Social Network Model

### 2019-01-25 ASApaper/paper-child.Rnw
some rearranging, re-write abstract.

\section{Introduction}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
Three of the most important pieces of statistical modeling are significance testing of model parameters, \st{model goodness-of-fit tests, and determining} power of a \st{hypothesis} test. In the first, the data are usually assumed to come from a simple model under the null hypothesis, and additional parameters are tested \st{to determine} whether they significantly contribute to explaining the variability in the data. In the second, the model of interest is examined to determine how well it fits the data. In the final, the ability of the hypothesis test in question to detect the difference between the null and alternative hypothesis is determined.  All three of these aspects of statistical modeling increase greatly in difficulty as the complexity of the model of interest increases. 
%The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

Some particularly complicated sets of models are those designed to model network change. A \textit{network} is any set of things, such as people, computers, or neurons, that are connected in some way, through social relations, internet connection, or electrical impulses in the brain. We refer to the ``things" in the network as \textit{nodes}, or \textit{actors} in a social network, and the connections as \textit{edges}, or \textit{ties} in a social network. 
Dependencies inherent to the data make network objects particularly difficult to model. %Even more challenging is the situation 
\st{THe difficulty of modeling increases} when we go beyond single instances of a network and consider the dynamics of network change between observed instances. This type of modelling for dynamic networks is often performed on social network data, such as friendship networks among students or the spread of HIV in drug users sharing needles. %sexually active communities.% of men sleeping with other men. X ST a bit less jarring hopefully. X 
%These 
\st{It is known that most network} models lack the asymptotics required to perform many well-known goodness-of-fit tests, and the maximum likelihood estimation of parameters is so difficult that it can make significance testing difficult as well \citep{goldenberg09}. 
%\hh{is the previous sentence a quote? I'm not sure that I follow.} %\st{it's a paraphrase... shall I make a quote? Also I definitely botched a word. Hopefully makes more sense now? }

\st{In order to circumvent these difficulties,} we propose new methods for significance \st{testing of parameters}, goodness-of-fit testing, and power calculations %of these 
for a set of social network models: stochastic actor-oriented models for dynamic network data \citep{saompaper}. Specifically, we are using \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald or $t$-tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. Visual inference, introduced by \citet{Bujaetal}, allows us to look at the \textit{entire} dataset simulated from a network model as opposed to a (set of) usually one-dimensional metric(s) derived from the network such as outdegree, or a $p$-value for a single parameter in the model. \st{By using visual inference to supplement traditional statistical tests, we gain a new understanding of parameters in SAOMs and fit of SAOMs to data.}%In addition, we use visual inference to better understand some parameters in a social network model, by looking at how different values for parameters affect the networks simulated from these models.

@@

\section{Visual Inference}\label{vi}

Data visualizations are an important component of data analysis, providing a mechanism for discovering patterns in data. Pioneering research by \citet{gelman:2004}, \citet{Bujaetal} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations. \citet{Bujaetal} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis. %The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model. The Rorschach protocol helps to understand the extent of randomness in the null model. 
Here, we use the lineup protocol. \st{Under this protocol, we begin with a data set of interest to us, such as a network, and a visualization of this data, such as a node-link diagram. We would like to know the model that generated this data. There are two hypothesis considered: a null hypothesis which states the model that is assumed to have generated the data, and an alternative hypothesis that the data were not generated under this model.} Then, the plot of the data of interest is placed randomly among a set of $m-1$ null plots, which are visualizations of data simulated from the null model(where $m =20$, usually). Human observers are then asked to examine the lineup and to identify the \st{plots that looks} most different \st{from the others}. If an observer identifies the data plot, this is quantifiable evidence against the null hypothesis. Since an observer has a chance of 1 in $m$ to pick the data plot from the lineup by simply guessing, i.e. in a situation where the data plot is virtually indistinguishable from the null plots, the evidence grows in strength with the number of independent observers identifying the data plot.

@@

The lineup protocol was formally tested in a head-to-head comparison with the equivalent conventional test in \citet{majumder:2011}. The experiment utilized human subjects from Amazon's Mechanical Turk \citep{turk} and used simulation to control conditions. The results suggest that  visual inference is comparable to conventional tests in a controlled conventional setting. This provides support for its appropriateness for testing in real exploratory situations where no conventional test exists or is difficult. %Interestingly, the power of a visual test increases with the number of observers engaged to evaluate lineups, and the pattern in results suggests that the power will provide results consistent with practical significance \citep{kirk:1996}.

\section{Stochastic Actor-Oriented Models}\label{sec:saom}

Broadly, a SAOM takes network structure and node covariate information into account in two ways and models the network changes \st{one-at-a-time} as a continuous time Markov chain (CTMC). First, the rate of change between states is dictated by a rate function that describes \textit{how often} changes in the network occur, and secondly, the objective function describes \textit{what} those state changes are. As in most other network models, the variables of interest are the edges of the network, $x_{ij}$, where $x_{ij}$ denotes the edge between nodes $i$ and $j$, where $i,j \in \{1, 2, \dots, n=\text{the number of nodes}\}$. 
$x_{ij}$ is modelled as a binary variable, i.e.
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
Edges are treated as \textit{directed}, i.e. in general $x_{ij} \neq x_{ji}$, and self-referencing edges or loops are not allowed, i.e.  $x_{ii} = 0$ for all $i$. 
The network is observed $M \geq 2$ times at time points  $t_1 < t_2 < ... < t_M$, and the entire network at time point $t_m$ is denoted as $x(t_m)$. In sections~\ref{rf} and~\ref{of} we discuss the rate and objective functions of a SOAM in more depth.  Additional details on SOAMs can be found in \citet{saompaper, snijders01, snijders:2010, snijders2010, snijdersetall:2007, snijders:2017},

\subsection{Rate Function}\label{rf}

All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network, i.e.\ each actor, $i$, gets a chance to make a change according to the rate function, typically denoted $\lambda_i$, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure e.g. outdegree of node $i$, and the node covariates into account, but we use the simple rate function, which is constant over all nodes in a given time period, so that $\lambda_i \equiv \alpha_m$ $\forall i \in \{1, 2, \dots, n\}$. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the \textit{waiting time} to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the \st{oppurtunity} to change \st{its set of ties} is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\subsection{Objective Function}\label{of}

After actor $i$ has been given the opportunity to change, it probabilistically chooses one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random component, $U$, which can be thought of as encompassing any other factors that may be influencing the changes the node makes not accounted for by the parameters in the model. %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ is aiming to maximize \st{its corresponding} objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$, given as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some network statistics, $s_{ik}(x, \mathbf{Z})$, $s_{ik}(x, \mathbf{Z})$, calculated with respect to actor $i$ \st{at the current network state $x$}. Network statistics range from the simple outdegree, $s_i(x) = \sum_{i\neq j} x_{ij}$, to the more complicated {\it transitive triplets jumping to different covariate}, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides  over 80 possible effects that can be included in the objective function. We discuss these statistics in more detail in Section~\ref{sec:models}.

This probability dictates which edge change is made by the acting node. The acting node can also choose to \textit{not} change at all. This \st{is most likely to occur} when the numerator, as calculated for the current state of the network, is larger than for any changes $x(i \leadsto j)$ that could be made.

According to \citet{RSienamanual}, at least two parameters must be included in the objective function: the density and the reciprocity. We denote the density, or out-degree, parameter by $\beta_1$ and the associated statistic as $s_{i1}(x) = \sum_{j} x_{ij}$. Similarly, we denote the reciprocity parameter by $\beta_2$ and the associated statistic as $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$. We \st{will} refer to the \st{very simple} model with only these two parameters in the objective function as M1.

\subsection{Example Data}\label{sec:senatedata}

The data we use are collaboration networks in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, overlapping with Barack Obama's presidency. These senates began on January 6, 2009, the start date of the $111^{th}$, and ended on January 3, 2017, the last date of the $114^{th}$\footnote{Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}.}. In the US Senate, there are three ways that senators can show support for a piece of legislation: they can author the bill, cosponsor the bill, and vote for the bill. We use cosponsorship as a metric because it results in a network that is unimodal (all nodes are senators) and directed. In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored. There are many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC). This value is defined in \citet{senate} as

@@

where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise. This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills. Because SAOMs require binary edges, we construct the edges as follows: 
 \begin{equation}\label{eq:edgewpc}
  x_{ij} =
\begin{cases}
                                   1 & WPC_{ij} > 0.25 \\
                                   0 & WPC_{ij} \leq 0.25
\end{cases}
\end{equation}
For each of the four senate sessions, \st{in addition to the} WPC value between any two senators in the session, we have three node covariates: the party affiliation of each senator, the number of bills they authored in each session, and their gender. We explored each of these covariates in the model to determine if they affect the overall network structure and how ties are formed between senators. The node-link diagram representations of the data we use for modelling are shown in Figure~\ref{fig:senateAll}. We have labelled some of the nodes in these networks whose names will be familiar to US readers, because they are leaders in their party or they have run for president. The size of the nodes represent how many bills the senator authored in a session, the color represents party affiliation, and the shape represent gender. In each of the four sessions, there is one very large connected component tying many of the prominent senators together, with many smaller groups of two to ten senators surrounding the larger component. In each senate, the structure changes slightly as new senators arrive or come to prominence.

@@

For Senate 111, for instance, we see Hillary Clinton, serving out her second term in the senate until she became Secretary of State. She is isolated in Figure~\ref{fig:senateAll}, but in actuality, she had many cosponsors on two pieces of legislation she authored in that short time, as is shown in Figure~\ref{fig:senateClinton}. We chose to remove Clinton and her edges from the network because they make the overall structure look so different from the other three senates, \st{where} the pattern is not typical of a senate in any other year. We suspect that because Hillary Clinton had \st{recently run for President had just} been \st{selected as Obama's} Secretary of State, the cosponsorships were largely symbolic, so the $111^{th}$ Senate without Hillary Clinton is more typical than the $111^{th}$ Senate with her.

<<senateClinton, fig.cap="We removed Hillary Clinton's ties from the network because she had abnormally high collaboration with senators during the time she was in the 111th senate and before she left office to become Secretary of State.", fig.height=3>>=
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)
clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name = "Bills\nauthored", range = c(.75, 3)) + 
  ThemeNet + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

In legislative cosponsorship networks, it is well known that party affiliation, reciprocity of relationships, and whether senators are "workhorses" who author many bills or "showhorses" who author few bills, are major influences on structure \citep{legnet}. We focus on these covariates, \st{plus the additional sex covariate} when choosing which SAO models to fit to the data.

\subsection{Models of Interest}\label{sec:models}

In addition to considering already well-known effects in legislative networks for application of our significance and goodness-of-fit methods, we first fit many other possible models and selected a few significant effects. To determine the effects that we would move forward with, we followed this procedure:

\begin{enumerate}
\item Define the simple effects structure of the data: the rate parameters and the outdegree and reciprocity parameters. 
\item Add each additional possible effect, \st{as determined by the effects documentation function}, in \texttt{RSiena} one-at-a-time to the model\st{'s objective function}  \citep{RSiena}.
\item Fit each model to the data and check for convergence.
    \begin{enumerate}
    \item If the model converged, move to 4.
    \item If the model did not converge, use the previous fitted values as starting values and repeat 5 times or until convergence, whichever comes first.
    \end{enumerate}
\item Test the added parameter for significance using a Wald-type test.
\item Report out the estimate of the additional parameter, its standard error, Wald $p$ value, and convergence criterion.
\end{enumerate}

After completing the procedure for all model effects, we selected effects whose estimates converged, had a Wald $p$-value of less than 0.10, and seemed to have a reasonable interpretation for our data according to well-known properties of legislative networks \citep{legnet}.

>>>>

\section{Background} \label{sec:bg}
% \st{
% \begin{enumerate}
% \item significance testing: what is it and why is it important
% \item goodness-of-fit testing: what is it and why is it important
% \item difficulty of these two for sna, why do vis inf
% \end{enumerate}
% }
When choosing and fitting statistical models, there are often three key evaluations of interest: significance testing of model parameters, model goodness-of-fit tests, and determining the power of those tests. With significance testing, the data are typically assumed to come from a simple model under the null hypothesis, and additional parameters are tested to determine whether they significantly contribute to explaining the variability in the data. In the second, the model of interest is examined to determine how well it fits the data. In the final, the ability of the hypothesis test in question to detect the difference between the null and alternative hypothesis is determined.  All three of these aspects of statistical modeling increase greatly in difficulty as the complexity of the model of interest increases. 
%The more complicated the model, the harder it is to determine fit or to include or exclude a parameter.

Some particularly complicated families of models are those designed to model network change. A \textit{network} is any set of things, such as people, computers, or neurons, that are connected in some way, through social relations, internet connection, or electrical impulses in the brain. We refer to the ``things" in the network as \textit{nodes}, or \textit{actors} in a social network, and the connections as \textit{edges}, or \textit{ties} in a social network. 
Dependencies inherent to the data make network objects particularly difficult to model. %Even more challenging is the situation 
The difficulty of modeling increases when we examine more than a single instance of a network and consider the dynamics of network change between observed instances. This type of modelling for dynamic networks is often performed on social network data, such as friendship networks among students or the spread of HIV in drug users sharing needles. %sexually active communities.% of men sleeping with other men. X ST a bit less jarring hopefully. X 
%These 
It is known that most network models lack the asymptotics required to perform many well-known goodness-of-fit tests, and the maximum likelihood estimation of parameters is so difficult that it can make significance testing difficult as well, see for example \citet{holland81}.
%\hh{is the previous sentence a quote? I'm not sure that I follow.} %\st{it's a paraphrase... shall I make a quote? Also I definitely botched a word. Hopefully makes more sense now? }

In order to circumvent some of these difficulties, we propose new methods for significance testing of parameters, goodness-of-fit testing, and power calculations %of these 
for a set of social network models: stochastic actor-oriented models for dynamic network data, first introduced by \citet{saompaper}. Specifically, we are using \textit{visual inference} in place of traditional statistical methods for social network models, such as Wald or $t$-tests for significance of parameters and in- and outdegree distribution metrics for determining goodness-of-fit. Visual inference, introduced by \citet{Bujaetal}, will allow us to look at the \textit{entire} dataset simulated from a network model as opposed to a (set of) usually one-dimensional metric(s) derived from the network such as outdegree distribution, or a $p$-value for a parameter in the model. By using visual inference to supplement traditional statistical tests, we gain a new understanding of parameters in SAOMs and fit of SAOMs to data.%In addition, we use visual inference to better understand some parameters in a social network model, by looking at how different values for parameters affect the networks simulated from these models.

@@

\subsection{Stochastic Actor-Oriented Models}\label{subsec:saom}

Broadly, a SAOM takes network structure and node covariate information into account in two ways and models the network changes one-at-a-time as a continuous time Markov chain (CTMC). First, the rate of change between states is dictated by a rate function that describes \textit{how often} changes in the network occur, and secondly, the objective function describes \textit{what} those state changes are. As in most other network models, the variables of interest are the edges of the network, $x_{ij}$, where $x_{ij}$ denotes the edge between nodes $i$ and $j$, where $i,j \in \{1, 2, \dots, n=\text{the number of nodes}\}$. 
$x_{ij}$ is modelled as a binary variable, i.e.
  \begin{equation}\label{eq:edgevars}
  x_{ij} =
  \begin{cases}
                                   1 & \text{if an edge from $i$ to $j$ exists} \\
                                   0 & \text{otherwise}
  \end{cases}
  \end{equation}
Edges are treated as \textit{directed}, i.e. in general $x_{ij} \neq x_{ji}$, and self-referencing edges or loops are not allowed, i.e.  $x_{ii} = 0$ for all $i=1, 2, \dots, n$. 
The network is observed $M \geq 2$ times at time points  $t_1 < t_2 < ... < t_M$, and the entire network at time point $t_m$ is denoted as $x(t_m)$. In sections~\ref{rf} and~\ref{of} we discuss the rate and objective functions of a SOAM in more depth.  Additional details on SOAMs can be found in \citet{saompaper, snijders01, snijders:2010, snijders2010, snijdersetall:2007, snijders:2017},

\subsubsection{Rate Function}\label{subsubsec:rf}

All changes in SAOMs are treated as changes made by the nodes, or \textit{actors}, in the network, i.e.\ each actor, $i$, gets a chance to make a change according to the rate function, typically denoted $\lambda_i$, which dictates when relationships between nodes in the network can change. In general, the rate function can take the network structure e.g. outdegree of node $i$, and the node covariates into account, but we use the simple rate function, which is constant over all nodes in a given time period, so that $\lambda_i \equiv \alpha_m$ $\forall i \in \{1, 2, \dots, n\}$. We denote the rate from $t_{m}$ to $t_{m+1}$ as $\alpha_m$ for $m = 1, \dots, M-1$. Using this notation, the \textit{waiting time} to the next chance for actor $i$ to make a change is exponentially distributed with expected value $\alpha_m^{-1}$. Since the rate is the same for all actors, the waiting time for \textit{any} actor to get the oppurtunity to change its set of ties is exponentially distributed with expected value $(n\alpha_m)^{-1}$.

\subsubsection{Objective Function}\label{subsubsec:of}

After actor $i$ has been given the opportunity to change, it probabilistically chooses one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random component, $U$, which can be thought of as encompassing any other factors that may be influencing the changes the node makes not accounted for by the parameters in the model. %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ is aiming to maximize its corresponding objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$, given as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some network statistics, $s_{ik}(x, \mathbf{Z})$, $s_{ik}(x, \mathbf{Z})$, calculated with respect to actor $i$ at the current network state $x$. Network statistics range from the simple outdegree, $s_i(x) = \sum_{i\neq j} x_{ij}$, to the more complicated {\it transitive triplets jumping to different covariate}, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides  over 80 possible effects that can be included in the objective function. We discuss these statistics in more detail in Section~\ref{sec:models}.

This probability dictates which edge change is made by the acting node. The acting node can also choose to \textit{not} change at all. This is most likely to occur when the numerator, as calculated for the current state of the network, is larger than for any changes $x(i \leadsto j)$ that could be made.

According to \citet{RSienamanual}, at least two parameters must be included in the objective function: the density and the reciprocity. We denote the density, or out-degree, parameter by $\beta_1$ and the associated statistic as $s_{i1}(x) = \sum_{j} x_{ij}$. Similarly, we denote the reciprocity parameter by $\beta_2$ and the associated statistic as $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$. We will refer to the very simple model with only these two parameters in the objective function as M1.

\subsection{Visual Inference}\label{subsec:vi}

Data visualizations are an important component of data analysis, providing a mechanism for discovering patterns in data. Pioneering research by \citet{gelman:2004}, \citet{Bujaetal} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations. \citet{Bujaetal} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis. %The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model. The Rorschach protocol helps to understand the extent of randomness in the null model. 
Here, we use the lineup protocol. Under this protocol, we begin with a data set of interest to us, such as a network, and a visualization of this data, such as a node-link diagram. We would like to know the model that generated this data. There are two hypothesis considered: a null hypothesis which states the model that is assumed to have generated the data, and an alternative hypothesis that the data were not generated under this model. Then, the plot of the data of interest is placed randomly among a set of $m-1$ null plots, which are visualizations of data simulated from the null model (where $m =20$, usually). Human observers are then asked to examine the lineup and to identify the plot(s) that look(s) most different from the others. If an observer identifies the data plot, this is quantifiable evidence against the null hypothesis. Since an observer has a chance of 1 in $m$ to pick the data plot from the lineup by simply guessing, i.e. in a situation where the data plot is virtually indistinguishable from the null plots, the evidence grows in strength with the number of independent observers identifying the data plot.

@@

The lineup protocol was formally tested in a head-to-head comparison with the equivalent conventional test in \citet{majumder:2011}. The experiment utilized human subjects from Amazon's Mechanical Turk \citep{turk} and used simulation to control conditions. The results suggest that  visual inference is comparable to conventional tests in a controlled conventional setting. This provides support for its appropriateness for testing in situations where no conventional test exists or is difficult. %Interestingly, the power of a visual test increases with the number of observers engaged to evaluate lineups, and the pattern in results suggests that the power will provide results consistent with practical significance \citep{kirk:1996}.

\section{Example Data and Models}\label{sec:senatedata}

\noindent \textbf{Data:} The data we use are collaboration networks in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, overlapping with Barack Obama's presidency. These senates began on January 6, 2009, the start date of the $111^{th}$ Congress, and ended on January 3, 2017, the last date of the $114^{th}$ Congress\footnote{Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}.}. In the US Senate, there are three ways that senators can show support for a piece of legislation: they can author the bill, cosponsor the bill, and vote for the bill. We use cosponsorship as a metric because it results in a network that is unimodal (all nodes are senators) and directed. In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored. There are many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC). This value is defined in \citet{senate} as

@@

where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise. This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills. Because SAOMs require binary edges, we construct the edges as follows: 
 \begin{equation}\label{eq:edgewpc}
  x_{ij} =
\begin{cases}
                                   1 & WPC_{ij} > 0.25 \\
                                   0 & WPC_{ij} \leq 0.25
\end{cases}
\end{equation}
For each of the four senate sessions, in addition to the WPC value between any two senators in the session, we have three node covariates: the party affiliation of each senator, the number of bills they authored in each session, and their gender. We explored each of these covariates in the model to determine if they affect the overall network structure and how ties are formed between senators. The node-link diagram representations of the data we use for modelling are shown in Figure~\ref{fig:senateAll}. We have labelled some of the nodes in these networks whose names will be familiar to US readers, because they are leaders in their party or they have run for president. The size of the nodes represent how many bills the senator authored in a session, the color represents party affiliation, and the shape represent gender. In each of the four sessions, there is one very large connected component tying many of the prominent senators together, with many smaller groups of two to ten senators surrounding the larger component. In each senate, the structure changes slightly as new senators arrive or come to prominence.

@@

For Senate 111, for instance, we see Hillary Clinton, serving out her second term in the senate until she became Secretary of State. She is isolated in Figure~\ref{fig:senateAll}, but in actuality, she had many cosponsors on only two pieces of legislation she authored in that short time.%, as is shown in Figure~\ref{fig:senateClinton}. 
We chose to remove Clinton and her edges from the network because they make the overall structure look vastly more connected than the other three senates. We suspect that because she had just recently run for President and had just been selected as Obama's Secretary of State, the cosponsorships she garnered between the start of the new congress and becoming Secretary of State were largely symbolic, so the $111^{th}$ Senate without Hillary Clinton is more typical than the $111^{th}$ Senate with her. In fact, she had several co-sponsors on both of her pieces of legislation, which meant she had the maximum WPC value of one with about a dozen other senators. Thus, though she is in the Senate for a time, subsequent analyses will not include her, and will instead include her replacement, Kirsten Gillibrand.

<<senateClinton, fig.cap="We removed Hillary Clinton's ties from the network because she had abnormally high collaboration with senators during the time she was in the 111th senate and before she left office to become Secretary of State.", fig.height=3, eval=FALSE>>=
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)
clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name = "Bills\nauthored", range = c(.75, 3)) + 
  ThemeNet + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

In legislative cosponsorship networks, it is well known that party affiliation, reciprocity of relationships, and whether senators are ``workhorses" who author many bills or ``showhorses" who author few bills, are major influences on structure \citep{legnet}. We focus on these covariates, plus the additional gender covariate when choosing which SAO models to fit to the data.

\noindent \textbf{Models:} In order to determine which models to examine, we first consider already well-known effects in legislative networks. We also selected some effects which were significant according to the built-in Wald-type tests in \texttt{RSiena} for application of our significance and goodness-of-fit methods. We examine a total of six models, each identified by its objective function: 
\begin{enumerate}
\item Model M1: $f_{i}(x, \boldsymbol{\beta}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x)$
\item Model M3: $f_{i}(x, \boldsymbol{\beta}, \mathbf{p}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x) + \beta_3 s_{i3}(x, \mathbf{p})$
\item Model M4: $f_{i}(x, \boldsymbol{\beta}, \mathbf{s}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x) + \beta_4 s_{i4}(x, \mathbf{s})$
\item Model M5: $f_{i}(x, \boldsymbol{\beta}, \mathbf{b}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x) + \beta_5 s_{i5}(x, \mathbf{b})$
\item Model M6: $f_{i}(x, \boldsymbol{\beta}, \mathbf{p}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x) + \beta_6 s_{i6}(x, \mathbf{p})$
\item Model M7: $f_{i}(x, \boldsymbol{\beta}, \mathbf{p}, \mathbf{b}, \mathbf{s}) = \beta_1 s_{i1}(x) + \beta_2 s_{i2}(x) + \beta_4 s_{i4}(x, \mathbf{s}) + \beta_5 s_{i5}(x, \mathbf{b}) + \beta_6 s_{i6}(x, \mathbf{p})$
\end{enumerate}

The parameters in the objective function, $\beta_1, \dots, \beta_6$ are defined in Table~\ref{tab:effects}.

%To determine the effects that we would move forward with, we followed this procedure:

% \begin{enumerate}
% \item Define the simple effects structure of the data: the rate parameters and the outdegree and reciprocity parameters. 
% \item Add each additional possible effect, \st{as determined by the effects documentation function}, in \texttt{RSiena} one-at-a-time to the model\st{'s objective function}  \citep{RSiena}.
% \item Fit each model to the data and check for convergence.
%     \begin{enumerate}
%     \item If the model converged, move to 4.
%     \item If the model did not converge, use the previous fitted values as starting values and repeat 5 times or until convergence, whichever comes first.
%     \end{enumerate}
% \item Test the added parameter for significance using a Wald-type test.
% \item Report out the estimate of the additional parameter, its standard error, Wald $p$ value, and convergence criterion.
% \end{enumerate}
%  
% After completing the procedure for all model effects, we selected effects whose estimates converged, had a Wald $p$-value of less than 0.10, and seemed to have a reasonable interpretation for our data according to well-known properties of legislative networks \citep{legnet}.
